Rényi Entropy for Heterogeneity Identification in SAR Data

### 2025-02-02 Text/Reports/report-Renyi/R1-report-renyi.Rmd
Add initial version of the paper

@@

>>>>

## Simplified Analytical Expressions for Rényi Entropy

Below we show how to express the Rényi entropy of the \( G_I^0 \) distribution in terms of the entropy of the \(\Gamma_{\text{SAR}}\) law plus additional terms involving \(\alpha\). We also ensure that the term \(\ln(\mu)\) remains explicitly isolated.

### 1) Rényi Entropy of \(\Gamma_{\text{SAR}}\)

For a \(\Gamma_{\text{SAR}}(L,\mu)\) random variable, a convenient form of its Rényi entropy of order \(\lambda\) (\(\lambda>0\), \(\lambda\neq 1\)) is:

\[
H_{\lambda}\bigl(\Gamma_{\text{SAR}}(L,\mu)\bigr)
\;=\;
\ln(\mu)\;+\;A_{\lambda}(L),
\tag{1}
\]

where

\[
A_{\lambda}(L)
\,=\,
-\,\ln L
\;+\;
\frac{\lambda}{\lambda \,-\,1}\,\ln\Gamma(L)
\;-\;
\frac{1}{\lambda \,-\,1}\,
\ln\Gamma\Bigl(\lambda\bigl(L - 1\bigr)+1\Bigr)
\;+\;
\frac{\lambda\bigl(L - 1\bigr)+1}{\lambda \,-\,1}
\,\ln(\lambda).
\]

Here, \(\Gamma(\cdot)\) is the gamma function. Crucially, notice that \(\ln(\mu)\) appears **as a separate term**.

### 2) Rényi Entropy of \( G_I^0 \)

For \( G_I^0(\alpha,L,\mu)\) (with \(\alpha<0\)), we can write its Rényi entropy in a comparable form:

\[
H_{\lambda}\bigl(G_I^0(\alpha,L,\mu)\bigr)
\;=\;
\ln(\mu)\;+\;A_{\lambda}(L)\;+\;T_{\lambda}(\alpha,L).
\tag{2}
\]

In this expression:

- \(\ln(\mu)\) is **exactly** the same term as in \(\Gamma_{\text{SAR}}\).  
- \(A_{\lambda}(L)\) is **the same function of \(L\) and \(\lambda\)** from Equation&nbsp;\((1)\).  
- \(T_{\lambda}(\alpha,L)\) **captures the extra terms** that arise due to the roughness parameter \(\alpha\).

Hence, by defining:

\[
T_{\lambda}(\alpha,L)
\;=\;
H_{\lambda}\bigl(G_I^0(\alpha,L,\mu)\bigr)
\;-\;
\Bigl[\,\ln(\mu)\,+\,A_{\lambda}(L)\Bigr],
\]

we isolate all the \(\alpha\)-dependent components in one place. In particular, one of the key additional factors is \(\ln(-1-\alpha)\), which does **not** appear in the \(\Gamma_{\text{SAR}}\) case. As \(\alpha\to -\infty\), these extra terms tend to the constants (or zero) that make the \(G_I^0\) law converge to \(\Gamma_{\text{SAR}}\).

### 3) Interpretation

1. **Homogeneous Case** (\(\Gamma_{\text{SAR}}\)):  
   \(\alpha\to -\infty\), thus \(T_{\lambda}(\alpha,L)\to 0\). Consequently,
   \[
   H_{\lambda}\bigl(G_I^0(\alpha\to -\infty,L,\mu)\bigr)
   \;=\;
   \ln(\mu) \;+\; A_{\lambda}(L),
   \]
   which exactly matches \(H_{\lambda}(\Gamma_{\text{SAR}})\).

2. **Heterogeneous Case** (\(G_I^0\), finite \(\alpha<0\)):  
   The additional term \(T_{\lambda}(\alpha,L)\neq 0\) captures the deviation from full speckle development, i.e., roughness. The more negative \(\alpha\) is, the closer \(G_I^0\) is to \(\Gamma_{\text{SAR}}\); the less negative \(\alpha\) is (e.g., \(\alpha=-3\)), the greater the discrepancy.

In this way, **\(H_{\lambda}(\Gamma_{\text{SAR}})\) acts as a baseline** for the fully developed speckle (homogeneous) scenario, and any departure from this baseline—quantified by \(T_{\lambda}(\alpha,L)\)—indicates the presence of heterogeneity.

## Simplified Analytical Expressions for Rényi Entropy

Below are *factorized* forms of the Rényi entropy for the two main models used in SAR image analysis, namely the \(\Gamma_{\text{SAR}}\) distribution (fully developed speckle) and the \(G_I^0\) distribution (able to represent roughness). These versions emphasize that \(\ln(\mu)\) appears **as an isolated term**, and they gather the remaining factors under a common \(\frac{1}{\lambda - 1}\) multiplier.

---

### 1) Rényi Entropy of \(\Gamma_{\text{SAR}}(L,\mu)\)

For a \(\Gamma_{\text{SAR}}(L,\mu)\) random variable, the Rényi entropy of order \(\lambda\) (\(\lambda>0, \lambda\neq1\)) can be written as:

\[
H_{\lambda}\bigl(\Gamma_{\text{SAR}}(L,\mu)\bigr)
=\;
\ln\bigl(\mu\bigr)
\;+\;
\frac{1}{\lambda - 1}\Bigl[
  \lambda\,\ln\Gamma(L)
  \;-\;
  \ln\Gamma\bigl(\lambda(L - 1)+1\bigr)
  \;-\;
  (\lambda - 1)\,\ln L
  \;+\;
  \bigl(\lambda(L - 1)+1\bigr)\,\ln\lambda
\Bigr].
\tag{1}
\]

Here, \(\Gamma(\cdot)\) is the gamma function. Notice how \(\ln(\mu)\) is an independent addend.

The bracketed term vanishes (or becomes a constant offset) as \(\alpha \to -\infty\), showing that \(\Gamma_{\text{SAR}}\) is indeed a limiting case of \(G_I^0\). In other words, when \(\alpha\) is very negative, the extra “roughness” contribution disappears, and the *fully developed speckle* scenario (homogeneous model) is recovered.

@@

\section*{Appendix: Limit Behavior of $H_\lambda(G_I^0)$ as $\alpha \to -\infty$}

\subsection*{Detailed Version of the Proof}

\begin{proof}
We want to show that 
\[
\lim_{\alpha \to -\infty}
H_\lambda\bigl(G_I^0\bigr)(\mu, \alpha, L)
=
H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)(\mu, L).
\]

First, recall that:
\begin{align}
H_\lambda\bigl(G_I^0\bigr)(\mu,\alpha,L)
&=
H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)(\mu,L)
\;+\;
\ln\!\bigl(-1-\alpha\bigr)
\\
&\quad
+ \frac{1}{1-\lambda}
\ln \Biggl[
  \frac{
    \Gamma(L-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(-\alpha+1)-1\bigr)\,\lambda^{\lambda(L-1)+1}
  }{
    \Gamma(-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(L-\alpha)\bigr)
  }
\Biggr].
\end{align}
Set
\begin{align}
\Delta(\alpha)
&=
H_\lambda\bigl(G_I^0\bigr)(\mu,\alpha,L)
-
H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)(\mu,L).
\end{align}
Then 
\[
\Delta(\alpha)
=
\ln(-1-\alpha)
+
\frac{1}{1-\lambda}
\ln \biggl[
  \frac{
    \Gamma(L-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(-\alpha+1)-1\bigr)\,\lambda^{\lambda(L-1)+1}
  }{
    \Gamma(-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(L-\alpha)\bigr)
  }
\biggr].
\]

As \(\alpha \to -\infty\), we have \(-1-\alpha \approx |\alpha|\), so 
\[
\ln(-1-\alpha) 
\sim 
\ln|\alpha|.
\]
Next, note that for large \(|\alpha|\), one uses the asymptotic relation 
\(
\Gamma(x+a)/\Gamma(x+b) 
\sim 
x^{\,a-b}
\).
Specifically:
\[
\Gamma(L-\alpha)/\Gamma(-\alpha) \;\sim\; |\alpha|^L,
\quad
\Gamma\bigl(\lambda(-\alpha+1)-1\bigr)/\Gamma\bigl(\lambda(L-\alpha)\bigr)
\;\sim\; 
\bigl(\lambda|\alpha|\bigr)^{\,(\lambda-1)-\lambda L}.
\]
Thus, inside the big logarithm,
\[
\frac{
  \Gamma(L-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(-\alpha+1)-1\bigr)
}{
  \Gamma(-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(L-\alpha)\bigr)
}
\;\sim\;
|\alpha|^{\lambda L}
\times
|\alpha|^{(\lambda-1)-\lambda L}
=
|\alpha|^{\,\lambda-1}.
\]
Multiplying by \(\lambda^{\lambda(L-1)+1}\) (a constant factor in \(\alpha\)) does not affect the \(\alpha\)-dependence. Therefore,
\[
\frac{1}{1-\lambda}\,
\ln\!\Bigl[\dots\Bigr]
\;\sim\;
\frac{1}{1-\lambda}\;\ln\!\bigl(|\alpha|^{\,\lambda-1}\bigr)
=
\frac{\lambda-1}{1-\lambda}\,\ln|\alpha|
=
-\ln|\alpha|.
\]
Hence
\[
\Delta(\alpha)
\;\sim\;
\ln|\alpha|
-
\ln|\alpha|
=
0
\quad
\text{as}
\;\;
\alpha\to -\infty.
\]
This shows 
\(\Delta(\alpha)\to 0\), 
and consequently
\[
\lim_{\alpha \to -\infty}
H_\lambda\bigl(G_I^0\bigr)(\mu,\alpha,L)
=
H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)(\mu,L).
\]
\end{proof}

\subsection*{Concise Version of the Proof}

\begin{proof}
We need to prove
\(
\lim_{\alpha\to -\infty} 
H_\lambda(G_I^0)(\mu, \alpha, L)
=
H_\lambda(\Gamma_{\mathrm{SAR}})(\mu, L).
\)
By the known decomposition,
\begin{align}
H_\lambda\bigl(G_I^0\bigr)(\mu,\alpha,L)
&=
H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)(\mu,L)
\;+\;
\ln\bigl(-1-\alpha\bigr)
\\
&\quad
+ \frac{1}{1-\lambda}
\ln \Biggl[
  \frac{
    \Gamma(L-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(-\alpha+1)-1\bigr)\,\lambda^{\lambda(L-1)+1}
  }{
    \Gamma(-\alpha)^{\lambda}\,\Gamma\bigl(\lambda(L-\alpha)\bigr)
  }
\Biggr].
\end{align}
Denote the extra \(\alpha\)-dependent terms by \(\Delta(\alpha)\).  As \(\alpha \to -\infty\), we have \(\ln(-1-\alpha)\sim\ln|\alpha|\).  Meanwhile, using 
\(\Gamma(x+a)/\Gamma(x+b)\sim x^{\,a-b}\),
the ratio of Gamma functions behaves like 
\(\,|\alpha|^{\lambda-1}\),
and multiplying by \(\lambda^{\lambda(L-1)+1}\) is a constant factor. Hence the logarithm of that ratio is \(\sim(\lambda-1)\ln|\alpha|\), and dividing by \((1-\lambda)\) gives \(-\ln|\alpha|\). Thus the two terms 
\(\ln|\alpha|\) 
and 
\(-\ln|\alpha|\) 
cancel each other. Therefore, 
\(\Delta(\alpha)\to 0\),
which concludes
\[
\lim_{\alpha \to -\infty} 
H_\lambda\bigl(G_I^0\bigr)(\mu,\alpha,L)
=
H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)(\mu,L).
\]
\end{proof}
